Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN normlaisation)

### 2017-04-06 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Minor modifications

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gc_content"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)

@@

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=8, fig.height=8}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
tmp = COVARIATES.CORRELATION$ESTIMATE
tmp[COVARIATES.CORRELATION$PVAL > 0.05] = 0
h = Heatmap(tmp, col = colorRamp2(c(-1,0,1), c('blue','white','red')), name = 'AssocEstimate')
draw(h, heatmap_legend_side = 'left')
```

@@

# Gender
# vcd::mosaic(~ Gender + Dx.Tissue, data = COVARIATES)
```
### Filter genes
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
genesToAnalyze = dlply(METADATA, .(Dx.Tissue), .fun = function(mtd, count){
  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID %>% droplevels()],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

@@

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples from `r length(unique(COVARIATES$Individual_ID))` unique individuals

### Outlier analysis
Detect outlier samples based on their expression (logCPM) pattern
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(voom(PROCESSED_COUNTS$fi)$E, scale.=T, center = T)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based classification of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Gender", "LibraryBatch", "Dx.Tissue")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

>>>>

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gc_content", "gene_biotype"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)

@@

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=10, fig.height=8}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
tmp = COVARIATES.CORRELATION$ESTIMATE
tmp[COVARIATES.CORRELATION$PVAL > 0.05] = 0
h = Heatmap(tmp, col = colorRamp2(c(-1,0,1), c('blue','white','red')), name = 'AssocEstimate')
draw(h, heatmap_legend_side = 'left')
```

@@

# Gender
# vcd::mosaic(~ Gender + Dx.Tissue, data = COVARIATES)
```
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
genesToAnalyze = dlply(METADATA, .(Dx.Tissue), .fun = function(mtd, count){
  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID %>% droplevels()],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

@@

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
pct.pc = PROCESSED_COUNTS$filteredExprMatrix$genes %>%
  dplyr::rename(ensembl_gene_id = genes) %>%
  left_join(Ensemble2HGNC) %>%
  group_by(gene_biotype) %>%
  summarise(fraction  = n()) %>%
  filter(fraction > 100) %>%
  dplyr::mutate(fraction = fraction/length(PROCESSED_COUNTS$filteredExprMatrix$genes[,1]))
```
`r dim(PROCESSED_COUNTS$filteredExprMatrix$genes)[1]` genes are used for the analysis.

Following is the distribution of biotypes
`r kable(pct.pc)`

### Outlier analysis
Detect outlier samples based on their expression (logCPM) pattern
```{r detect.outliers, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E, scale.=T, center = T)

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
  dplyr::mutate(label = SampleID)
plotdata$label[!(plotdata$SampleID %in% c('MSSM_RNA_ACC_BP_24', 
                                          'MSSM_RNA_ACC_375'))] = ''

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
p <- p + geom_text(aes(label= label), size=4, hjust=0)
p

# Plot abberent distribution of logcpm counts
tmp1 = voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  tidyr::gather(SampleID, logCPM, -ensembl_gene_id) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID') %>%
              tidyr::separate(Dx.Tissue, c('Dx', 'Tissue'), sep = '_'))

p = ggplot(tmp1 %>%
             dplyr::filter(SampleID %in% c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375')),
           aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'top') + facet_grid(Dx+.~Tissue, scale = 'free')
p

indToRemove = c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375')
indToRetain = setdiff(colnames(PROCESSED_COUNTS$filteredExprMatrix$counts), indToRemove)
PROCESSED_COUNTS$filteredExprMatrix$counts = PROCESSED_COUNTS$filteredExprMatrix$counts[,indToRetain]
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples from `r length(unique(COVARIATES$Individual_ID))` unique individuals

Based on the expression pattern following samples were tagged as outliers: `r paste(indToRemove, collapse = ', ')`

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p

# Plot abberent distribution of logcpm counts
tmp1 = (CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset) %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  tidyr::gather(SampleID, logCPM, -ensembl_gene_id) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID') %>%
              tidyr::separate(Dx.Tissue, c('Dx', 'Tissue'), sep = '_'))

p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'NONE') + facet_grid(Dx+.~Tissue, scale = 'free')
p
```
Tree based classification of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Gender", "LibraryBatch", "Dx.Tissue")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
